A formal treatment of bidirectional typing

### 2024-01-11 tex/ESOP24/BiSig.tex
[ tex ] Some revisions before removing appendix for the camera ready version

\author{Liang-Ting Chen\orcidID{0000-0002-3250-1331} \and Hsiang-Shang Ko\orcidID{0000-0002-2439-1048}}
\institute{Institute of Information Science, Academia Sinica, Taiwan}
\title{A formal treatment of bidirectional typing\texorpdfstring{\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}{}}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Sufficient conditions for deriving a type synthesiser such as soundness, completeness, and mode-correctness, are studied universally for all signatures.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

>>>>

\author{Liang-Ting Chen\orcidID{0000-0002-3250-1331} \and Hsiang-Shang Ko\orcidID{0000-0002-2439-1048}}
\institute{Institute of Information Science, Academia Sinica, Taiwan}
\title{A formal treatment of bidirectional typing}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a \emph{proof-relevant} type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Sufficient conditions for deriving a type synthesiser such as soundness, completeness, and mode-correctness, are studied universally for all signatures.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a raw term.
The entire theory is formally implemented in \Agda, so as a by-product of our formalism we provide a \emph{verified} generator of proof-relevant type synthesisers.
\end{abstract}

### 2024-01-11 tex/ESOP24/S1-intro.tex
[ tex ] Some revisions before removing appendix for the camera ready version

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
In their survey paper, \varcitet{Dunfield2021}{ summarised} the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
For example, the type of a variable can be looked up in the context.
Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
Formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is logical and computational, a decidability proof is a proof-relevant decision procedure that computes a yes-or-no answer with a proof of~$P$ or its refutation, so logical decidability is algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (implying the original form due to the decidability).
Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also its justification.
This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are loosely related.

@@

As mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
In our theory, we define \emph{mode derivations} so that we can explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.

The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
Therefore, to solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provides information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.

@@

In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the typed version of \varcitet{Aczel1978}{'s binding signature} with modes.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).

To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
It should be possible to extend the theory to deal with more expressive types and assumptions other than mode-correctness.
For instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.

>>>>

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism entails undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, deciding the types of raw terms that meet some syntactic criteria and typically contain annotations.
In their survey paper, \varcitet{Dunfield2021}{ summarised} the principles of bidirectional type synthesis and its wide coverage of languages with simple, polymorphic, dependent, and gradual types, among others.

While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
For example, the type of a variable can be looked up in the context.
Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
Formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can often be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is logical and computational, a decidability proof is a proof-relevant decision procedure that computes a yes-or-no answer with a proof of~$P$ or its refutation, so logical decidability is algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
Therefore, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are loosely related.

@@

As mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
In our theory, we define \emph{mode derivations} to explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.

The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
Thus, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide if the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a proof-relevant \emph{mode decorator}, which either
\begin{inlineenum}
  \item construct a mode derivation for a raw term, or
  \item provides information that refutes the existence of any mode derivation and pinpoints missing annotations.
\end{inlineenum}
Then a bidirectional type synthesiser is only required to decide the typability of mode-decorated raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.

@@

In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the typed version of \varcitet{Aczel1978}{'s binding signature} with modes.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for mode-correct signatures.
The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).

To make our exposition accessible, the theory in this paper focuses on simply typed languages with a \emph{syntax-directed} bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
It should be possible to extend the theory to deal with more expressive types and assumptions other than mode-correctness.
For instance, we briefly discuss how the theory could be extended to handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.

### 2024-01-11 tex/ESOP24/S2-outline.tex
[ tex ] Rewrite the example for binding arity with colourful annotations
[ tex ] Some revisions before removing appendix for the camera ready version

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
To give more precise definitions:
The raw terms for simply typed $\lambda$-calculus are defined%
\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.

\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{V |- \isTerm{t}}$ & Given a list~$V$ of variables, $t$~is a raw term with free variables in~$V$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{x \in V}{V \vdash \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{V \vdash \isTerm{t}}{V \vdash \isTerm{(t \annotate A)}}\,\Rule{Anno}
    \and
    \inferrule{V, \isTerm{x} \vdash \isTerm{t}}{V \vdash \isTerm{\lam{x}t}}\,\Rule{Abs}
    \and
    \inferrule{V \vdash \isTerm{t} \\ V \vdash \isTerm{u}}{V \vdash \isTerm{t\;u}}\,\Rule{App}
  \end{mathpar}
  \caption{Raw terms for simply typed $\lambda$-calculus}
  \label{fig:STLC-raw-terms}
\end{figure}

\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{\Gamma |- \isTerm{t} : A}$ & A raw term~$t$ has type~$A$ under context~$\Gamma$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} : A}\,\Rule{Var}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash \isTerm{(t \annotate A)} : A}\,\Rule{Anno}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} : B}{\Gamma \vdash \isTerm{\lam{x}t} : A \bto B}\,\Rule{Abs}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A \bto B \\ \Gamma \vdash \isTerm{u} : A}{\Gamma \vdash \isTerm{t\;u} : B}\,\Rule{App}
  \end{mathpar}
  \caption{Typing derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-typing-derivations}
\end{figure}
Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
%\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
Now we can define what it means to solve the type synthesis problem.

@@

It is crucial to allow the third outcome, without which we would be requiring the type synthesis problem to be decidable, but this requirement would quickly become impossible to meet when the theory is extended to handle more complex types.
If a type synthesiser cannot decide whether there is a typing derivation, it is allowed to give an excuse instead of an answer.
Acceptable excuses are defined by the predicate~$E$, which describes what is wrong with an input term, for example not having enough type annotations.

\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{\Gamma |- \isTerm{t} \syn A}$ & A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
    $\boxed{\Gamma |- \isTerm{t} \chk A}$ &A raw term $t$ checks against a type $A$ under $\Gamma$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \syn A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{(t \annotate A)}\syn A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ B = A}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn A \bto B \\ \Gamma \vdash \isTerm{u} \chk A}{\Gamma \vdash \isTerm{t\;u} \syn B}\,\SynRule{App}
  \end{mathpar}
  \caption{Bidirectional typing derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-bidirectional-typing-derivations}
\end{figure}

Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
It is often said that a type synthesis algorithm can be `read off' from well-designed bidirectional typing rules.
Take the $\SynRule{App}$ rule as an example:
to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the type $A \bto B$.
Note that the synthesiser is able to figure out the type~$A$ for checking~$u$ and the type~$B$ to be synthesised for $t\;u$ because they have been computed when synthesising the type $A \bto B$ of~$t$.
In general, there should be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised).
This is called \emph{mode-correctness}, which we will formally define in \cref{sec:mode-correctness}.

@@

\begin{figure}
  \centering
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{V |- \isTerm{t}^{\syn}}$ & A raw term~$\isTerm{t}$ (with free variables in~$V$) is in synthesising mode \\
    $\boxed{V |- \isTerm{t}^{\chk}}$ & A raw term~$\isTerm{t}$ (with free variables in~$V$) is in checking mode
  \end{tabular}
  \egroup
  \begin{mathpar}
    \inferrule{x \in V}{V |- \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{V |- \isTerm{t}^{\chk}}{V |- (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{V |- \isTerm{t}^{\syn}}{V |- \isTerm{t}^{\chk}}\,\ChkRule{Sub}
    \\
    \inferrule{V, x |- \isTerm{t}^{\chk}}
    {V |- (\isTerm{\lam{x}t})^{\chk}}\,\ChkRule{Abs}
    \and
    \inferrule{V |- \isTerm{t}^{\syn} \\ V |- \isTerm{u}^{\chk}}
    {V |- (\isTerm{t\;u})^{\syn}}\,\SynRule{App}
  \end{mathpar}
  \caption{Mode derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-mode-derivations}
\end{figure}

@@

\begin{defn} \label{def:STLC-mode-decorator}
A \emph{mode decorator} decides for any raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
\end{defn}

@@

Now we can get back to our goal of implementing a type synthesiser~(\cref{def:STLC-type-synthesiser}).

@@

The construction is straightforward:
run the mode decorator on the input term~$\erase\Gamma \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}); if there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, then there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete}:
\begin{theorem}[Soundness and Completeness]
$\Gamma \vdash t \syn A$ if and only if\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$.
\end{theorem}
%
%
%\begin{lemma}[Completeness] \label{thm:STLC-completeness}
%If\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
%\end{lemma}

>>>>

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take a raw term as input, and produce a typing derivation for the term if possible.
To give more precise definitions: the raw terms for simply typed $\lambda$-calculus are defined%
\footnote{The usual condition about named representations of variables is omitted.}
in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.

\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{V |- \isTerm{t}}$ & Given a list~$V$ of variables, $t$~is a raw term with free variables in~$V$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{x \in V}{V \vdash \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{V \vdash \isTerm{t}}{V \vdash \isTerm{(t \annotatecolon A)}}\,\Rule{Anno}
    \and
    \inferrule{V, \isTerm{x} \vdash \isTerm{t}}{V \vdash \isTerm{\lam{x}t}}\,\Rule{Abs}
    \and
    \inferrule{V \vdash \isTerm{t} \\ V \vdash \isTerm{u}}{V \vdash \isTerm{t\;u}}\,\Rule{App}
  \end{mathpar}
  \caption{Raw terms for simply typed $\lambda$-calculus}
  \label{fig:STLC-raw-terms}
\end{figure}

\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{\Gamma |- \isTerm{t} : A}$ & A raw term~$t$ has type~$A$ under context~$\Gamma$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} : A}\,\Rule{Var}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash \isTerm{(t \annotatecolon A)} : A}\,\Rule{Anno}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} : B}{\Gamma \vdash \isTerm{\lam{x}t} : A \bto B}\,\Rule{Abs}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A \bto B \\ \Gamma \vdash \isTerm{u} : A}{\Gamma \vdash \isTerm{t\;u} : B}\,\Rule{App}
  \end{mathpar}
  \caption{Typing derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-typing-derivations}
\end{figure}

Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of the simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} has an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.

%\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
Now we can define what it means to solve the type synthesis problem.

@@

It is crucial to allow the third outcome, without which we would be requiring the type synthesis problem to be decidable, but this requirement would quickly become impossible to meet when the theory is extended to handle more complex types.
If a type synthesiser cannot decide whether there is a typing derivation, it is allowed to give an excuse instead of an answer.
Acceptable excuses are defined by the predicate~$E$, which describes what is wrong with an input term, for example, not having enough type annotations.

\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{\Gamma |- \isTerm{t} \syn A}$ & A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
    $\boxed{\Gamma |- \isTerm{t} \chk A}$ &A raw term $t$ checks against a type $A$ under $\Gamma$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \syn A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{(t \annotatecolon A)}\syn A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ B = A}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn A \bto B \\ \Gamma \vdash \isTerm{u} \chk A}{\Gamma \vdash \isTerm{t\;u} \syn B}\,\SynRule{App}
  \end{mathpar}
  \caption{Bidirectional typing derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-bidirectional-typing-derivations}
\end{figure}

\begin{figure}
  \centering
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{V |- \isTerm{t}^{\syn}}$ & A raw term~$\isTerm{t}$ (with free variables in~$V$) is in synthesising mode \\
    $\boxed{V |- \isTerm{t}^{\chk}}$ & A raw term~$\isTerm{t}$ (with free variables in~$V$) is in checking mode
  \end{tabular}
  \egroup
  \begin{mathpar}
    \inferrule{x \in V}{V |- \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{V |- \isTerm{t}^{\chk}}{V |- (\isTerm{t \annotatecolon A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{V |- \isTerm{t}^{\syn}}{V |- \isTerm{t}^{\chk}}\,\ChkRule{Sub}
    \\
    \inferrule{V, x |- \isTerm{t}^{\chk}}
    {V |- (\isTerm{\lam{x}t})^{\chk}}\,\ChkRule{Abs}
    \and
    \inferrule{V |- \isTerm{t}^{\syn} \\ V |- \isTerm{u}^{\chk}}
    {V |- (\isTerm{t\;u})^{\syn}}\,\SynRule{App}
  \end{mathpar}
  \caption{Mode derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-mode-derivations}
\end{figure}
Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
It is often said that a type synthesis algorithm can be `read off' from well-designed bidirectional typing rules.
Take the $\SynRule{App}$ rule as an example:
to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the type $A \bto B$.
Note that the synthesiser is able to figure out the type~$A$ for checking~$u$ and the type~$B$ to be synthesised for $t\;u$ because they have been computed when synthesising the type $A \bto B$ of~$t$.
In general, there should be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised).
This is called \emph{mode-correctness}, which we will formally define in \cref{sec:mode-correctness}.

@@

@@

\begin{defn} \label{def:STLC-mode-decorator}
A \emph{mode decorator} decides for any raw term~$V \vdash \isTerm{t}$ if $V \vdash \isTerm{t}^{\syn}$.
\end{defn}

@@

Now we can get back to implementing a type synthesiser~(\cref{def:STLC-type-synthesiser}).

@@

The construction is straightforward:
run the mode decorator on the input term~$\erase\Gamma \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}); if there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, then there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete}:
\begin{theorem}[Soundness and Completeness]
$\Gamma \vdash t \syn A$ if and only if $\erase\Gamma \vdash t^{\syn}$ and $\Gamma \vdash t : A$.
\end{theorem}
%
%
%\begin{lemma}[Completeness] \label{thm:STLC-completeness}
%If\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
%\end{lemma}
